Improving Dark Matter Axion Searches with Active Resonators

### 2015-04-04 paper/paper.tex
Added Gertsenshtein, Tetu, Brida to bib file, jotted down thoughts about gain instability
Noise_Curves notebook showing theoretical SNR enhancement

\documentclass[aps,prl,twocolumn,groupedaddress]{revtex4-1}

%\documentclass[11pt]{amsart}
%\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amsmath}
%\usepackage{amssymb}
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\begin{document}

@@

Let us look at the amplitude of the wave at the output of port 2, assuming that all components are impedance matched and that isolators are placed at the ports to prevent reflected waves. If port 2 is critically coupled, then the noise wave at its output is equal to $c_2^2 =kT_{cav}B$, where B is the bandwidth of the detector.(G is the voltage gain of the amplifier and n_a the noise wave associated with it.)

The time averaged noise power of the output is then
\begin{equation}
P_n = <|(c_2(t) + n_a(t)) + (c_2(t-t_1)+n_a(t-t_1))GS_{21} + (c_2(t-2t_1)+n_a(t-2t_1))G^2S_{21}^2 + \ldots|^2>
\end{equation}
The amplifier noise $n_a$ and the thermal noise from the cavity $c_2$ are independent, hence $<c_2n_a> = 0$, and also in the steady state $<c_2(t)>^2 = \bar{c_2}$ is independent of time, so

\begin{equation}
P_n = (\bar{c_2}^2 + \bar{n_a}^2)\frac{1}{1 - G^2|S_{21}|^2} + \text{cross-terms} \\
\end{equation}
where the cross-terms hold the information about the coherence of the noise between feedback rounds. Any signal entering the cavity undergoes exponential decay, so the expectation value of the cross terms goes as $<(c_2(t-nt_1)c_2(t-mt_1)> = \bar{c_2}^2e^(-(|n-m|)t_1/\tau)$, where $\tau = Q_L/\omega_0$ is the coherence time of the cavity. One can see that when the roundtrip time around the feedback loop is much larger than the coherence time, $t_1 \gg \tau$, the cross terms are exponentially suppressed and the feedback loop and amplifier enhance the original noise temperature of the cavity by a factor:
\begin{equation}
T_n = T_{cav}[1+\frac{T_a}{T_{cav}\frac{1}{1-G^2|S_{21}|^2}
\end{equation}

The analysis is slightly different for the signal power. We treat the axion-converted signal as a continuous source with amplitude $e_a$. Then the signal power observed at the output of port 2 (with the feedback loop is
\begin{equation}
P_s = <|e_a(t) + e_a(t-t_1)GS_{21} + e_a(t-2t_1)G^2S_{21}^2 + \ldots|^2>
\end{equation}

and as long as \omega t_1 is a multiple of $2\pi$, $e_a(t-nt_1) = e_a(t)$.

Therefore the time averaged signal power is
\begin{equation}
P_s = \bar{e_a}^2\frac{1}{(1 - GS_{21})^2}
\end{equation}

@@

\begin{equation}
SNR_{e}/SNR_{0} =P_{s, enhanced}/P_{s,original}\frac{(1-G^2|S_{21}|^2}{(1-GS_{21})^2} = 1/(1-GS_{21})(1 + \frac{2GS_{21}}{1-GS_{21}}
\end{equation}
In most cases we can achieve loaded Q's of only an order of magnitude below the axion linewidth, so we need a combined feedback of $GS_{21} = .8$ to improve the Q by an order of magnitude.

@@

\begin{equation}
P_n = (\bar{c_2}^2 + \bar{n_a}^2)\frac{1}{1 - G^2|S_{21}|^2}\frac{1}{1-2e^{-t/\tau}GS_{21}}
\end{equation}

@@

As well, the enhanced Q (again assuming a phase shift of $2\pi$), is given by \cite{ref:frenchpaper}
\begin{equation}
Q_e = Q_L\frac{1}{1-GS_{21}}
\end{equation}
you can see this by noting that the Q is a measure of how much a signal builds up; since the signal builds up by 
$Q_{unloaded} = \omega U / P_{walls}$.
Similarly
$Q_{e1} = \omega U / P_{port1}$ and $Q_{e2} = \omega U / P_{port2}$.
Therefore the loaded passive Q is $Q_L^{-1} = P_{total lost}/\omega U =  = Q_{e1}^{-1} + Q_{e2}^{-2} + Q_{unloaded}^{-1}$.
Then with the feedback loop some power is regained, so 
$Q_e^{-1} = 1/\omega U (P_{port1} + P_{port2} + P_{walls} - P_{feedback}) = Q_L$
%% I can now treat the cavity as a three port device using amplitudes defined in Fig.~\ref{fig:amplitude_defs} with an $S$ matrix of
%% \begin{equation}
%% S = \left( \begin{array}{ccc}
%%   -1 + 2 \beta_1 (1 + \beta_1 + \beta_2)^{-1} & 2\sqrt{\beta_1\beta_2}(1 + \beta_1 + \beta_2)^{-1} \\
%%   2\sqrt{\beta_1\beta_2}(1 + \beta_1 + \beta_2)^{-1} &   -1 + 2 \beta_2 (1 + \beta_1 + \beta_2)^{-1} \\
%%   \end{array}\right)
%% \end{equation}

@@

We see in this prototype that with a very low Q cavity one can enhance the Q by a factor of more than a thousand, and see a signal to noise improvement of a factor of 10. This is extremely encouraging as a proof of principle, and shows that one can get to Q's comparable with the axion linewidth with the simple techniqure of active feedback. This will increase the scan rate of microwave cavity experiments roughly by a factor of 3 and is simpler than research currently under way to improve the Q by making superconducting films on the walls of copper cavities. The active feedback technique becomes increasingly important for high-frequency axion experiments, as the passive limit for the Q scales as $f^{-1/3}$.

@@

\bibliography{2014_ouroboros}

>>>>

\documentclass[aps,prl,twocolumn,groupedaddress]{revtex4-1}
%\documentclass[11pt]{amsart}
%\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amsmath}
%\usepackage{amssymb}
%\usepackage{epstopdf}
%\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\begin{document}

@@

Let us look at the amplitude of the wave at the output of port 2, assuming that all components are impedance matched and that isolators are placed at the ports to prevent reflected waves. If port 2 is critically coupled, then the noise wave at its output is equal to $c_2^2 =kT_{cav}B$, where B is the bandwidth of the detector.(G is the voltage gain of the amplifier and $n_a$ the noise wave associated with it.) We can add the noise of the cavity and the amplifier to get the system noise wave, $n(t) = c_2(t)+n_a(t)$.

The time averaged noise power of the output is then
\begin{equation}
P_n = \langle |n(t) + n(t-t_1)GS_{21} + (c_2(t-2t_1)+n_a(t-2t_1))G^2S_{21}^2 + \ldots|^2 \rangle
\end{equation}
The amplifier noise $n_a$ and the thermal noise from the cavity $c_2$ are independent, hence $<c_2n_a> = 0$, and also in the steady state $<c_2(t)>^2 = \bar{c_2}$ is independent of time, so

\begin{equation}
P_n = (\bar{c}_2^2 + \bar{n}_a^2)\frac{1}{1 - G^2|S_{21}|^2} + \text{cross-terms} \\
\end{equation}
where the cross-terms hold the information about the coherence of the noise between feedback rounds. Any signal entering the cavity undergoes exponential decay, so the expectation value of the cross terms goes as $<(c_2(t-nt_1)c_2(t-mt_1)> = \bar{c_2}^2e^(-(|n-m|)t_1/\tau)$, where $\tau = Q_L/\omega_0$ is the coherence time of the cavity. One can see that when the roundtrip time around the feedback loop is much larger than the coherence time, $t_1 \gg \tau$, the cross terms are exponentially suppressed and the feedback loop and amplifier enhance the original noise temperature of the cavity by a factor:
\begin{equation}
T_n = T_{sys}\frac{1}{1-G^2|S_{21}|^2}
\end{equation}

The analysis is slightly different for the signal power. We treat the axion-converted signal as a continuous source with amplitude $e_a$. Then the signal power observed at the output of port 2 (with the feedback loop is
\begin{equation}
P_s = \langle |e_a(t) + e_a(t-t_1)GS_{21} + e_a(t-2t_1)G^2S_{21}^2 + \ldots|^2 \rangle
\end{equation}

and as long as $\omega t_1$ is a multiple of $2\pi$, $e_a(t-nt_1) = e_a(t)$.

Therefore the time averaged signal power is
\begin{equation}
P_s = \bar{e}_a^2\frac{1}{(1 - GS_{21})^2}
\end{equation}

@@

\begin{align}
\frac{SNR_{e}}{SNR_{0}} &=\frac{P_{s, enhanced}}{P_{s,original}}\frac{T_o}{T_e} = \frac{Q_e}{Q_o}\frac{1-G^2|S_{21}|^2}{(1-GS_{21})^2} 
\\ &= \frac{1}{1-GS_{21}}\bigg(1 + \frac{2GS_{21}}{1-GS_{21}}\bigg)
\end{align}
In most cases we can achieve loaded Q's of only an order of magnitude below the axion linewidth, so we need a combined feedback of $GS_{21} = .8$ to improve the Q by an order of magnitude.

@@

\begin{equation}
P_n = (\bar{c}_2^2 + \bar{n}_a^2)\frac{1}{1 - G^2|S_{21}|^2}\frac{1}{1-2e^{-t/\tau}GS_{21}}
\end{equation}

@@

The SNR will plateau once we can no longer assume that the signal persists indefinitely - that is to say, when we include the $Q$ of the signal in our description of the signal power, we get that the
As well, the enhanced Q (again assuming a phase shift of $2\pi$), is given by \cite{ref:frenchpaper}
\begin{equation}
Q_e = Q_L\frac{1}{1-GS_{21}}
\end{equation}
you can see this by noting that the Q is a measure of how much a signal builds up; since the signal builds up by 
$Q_{unloaded} = \omega U / P_{walls}$.
Similarly
$Q_{e1} = \omega U / P_{port1}$ and $Q_{e2} = \omega U / P_{port2}$.
Therefore the loaded passive Q is $Q_L^{-1} = P_{total lost}/\omega U =  = Q_{e1}^{-1} + Q_{e2}^{-2} + Q_{unloaded}^{-1}$.
Then with the feedback loop some power is regained, so 
$Q_e^{-1} = 1/\omega U (P_{port1} + P_{port2} + P_{walls} - P_{feedback}) = Q_L$
%% I can now treat the cavity as a three port device using amplitudes defined in Fig.~\ref{fig:amplitude_defs} with an $S$ matrix of
%% \begin{equation}
%% S = \left( \begin{array}{ccc}
%%   -1 + 2 \beta_1 (1 + \beta_1 + \beta_2)^{-1} & 2\sqrt{\beta_1\beta_2}(1 + \beta_1 + \beta_2)^{-1} \\
%%   2\sqrt{\beta_1\beta_2}(1 + \beta_1 + \beta_2)^{-1} &   -1 + 2 \beta_2 (1 + \beta_1 + \beta_2)^{-1} \\
%%   \end{array}\right)
%% \end{equation}

@@

We see in this prototype that with a very low Q cavity one can enhance the Q by a factor of more than a thousand, and see a signal to noise improvement of a factor of 10. This is extremely encouraging as a proof of principle, and shows that one can get to Q's comparable with the axion linewidth with the simple techniqure of active feedback. The main sources of systematic noise would come from gain instability in the circuit; in this experiment no noise additional to that expected was observed (or tested). This will increase the scan rate of microwave cavity experiments roughly by a factor of 3 and is simpler than research currently under way to improve the Q by making superconducting films on the walls of copper cavities. The active feedback technique becomes increasingly important for high-frequency axion experiments, as the passive limit for the Q scales as $f^{-1/3}$.

@@

\bibliography{2015_ouroboros}
